# Setting Up Your RStudio Environment {#sec-setting-up-your-rstudio-environment}

We can go ahead and start building our time series analysis task right away after installing and loading the needed packages. However, to keep things organised you might want to;

1.  **Create a new project** in RStudio (`File > New > Project`)

2.  **Set your working directory** to the folder on your PC where your datasets are stored.

    ```{r}
    #| eval: false
    setwd("path/to/folder")

    ```

    ::: callout-tip
    When creating a new project you can directly set your project directory to your preferred directory by choosing **`"Existing Directory"`** and selecting the directory manually.
    :::

3.  Keep all scripts, data, and outputs in separate folders for better workflow

## Quick Test Run

Let's run a quick test to ensure everything is working. We shall use a built in dataset

```{r}
#| eval: false

# Use the 'us_employment' dataset from the fpp3 package
library(fpp3)
head(us_employment)

# Lets see if it is a proper tsibble and then we can plot it
us_employment |> 
  filter(Title == "Total Private") |> 
  autoplot(Employed) +
  labs(title = "US Total Private Employment",
       y = "People")
```

This "`head(us_employment)` " line of code should show you a tidy time series dataset with **index (time column)** and **key (group identifier)**. The codes that follow, if it runs without errors, should also produce a time series plot.

If you see all the things mentioned earlier, congratulations! Your environment is ready.

::: callout-note
The `fpp3` package is a collection of tools and datasets (which we have already seen) for time series forecasting. It was built to accompany the third edition of the book [***Forecasting: Principles and Practice***](https://otexts.com/fpp3/) by *`Rob J Hyndman and George Athanasopoulos`* .
:::
