# Required Packages {#sec-required-packages}

As stated earlier we will be using a combination of `tidyverse` and time series specific packages (`tidyverts`)

| Package | Purpose |
|------------------------------------|------------------------------------|
| `tidyverse` | Core suite for data wrangling and visualisation (`dplyr`, `tibble`, `ggplot2`, `readr`, etc) |
| `tsibble` | Tidy data frames for time series, handles dates, keys and indexing |
| `fable` | Forecasting models (ETS, ARIMA, Naive, etc) with tidy outputs |
| `feasts` | **F**eature **E**xtraction and **A**nalysis for **S**eries **T**ime **S**eries (plots, decomposition, autocorrelation, etc) |
| `lubridate` | Easy and readable date time manipulation |
| `readr` | Fast and easy data imports (csv, text data, etc) |
| `readxl` | Importing excel files (could be optional but very useful) |

There is another supporting package in the **tidyverts** ecosystem caled `fabletools`. This is a supporting package that provides some underlying tools for the `fables` package

::: callout-note
the `readr` and `lubridate` packages are already part of the **tidyverse package collections**. Once you install `tidyverse`, you do not need to install them again separately. `fabletools` is installed together with the `fables` package.
:::

## Installing the Packages

To install the above packages, run the following code **once** in your R console:

```{r}
#| eval: false

install.packages(c("tidyverse", "tsibble", "fable", "feasts", "readxl"))

```

::: callout-tip
If you already have them installed, you do not need to reinstall. You can check if a package is installed by typing `"fable" %in% installed.packages()`
:::

### Loading the Packages

After installing any package you have to load it before it becomes available for use. Each time you start a new R session, you will need to load the packages you will use again. Doing it correctly saves headaches. You can load a package in R with the `library()` function.

```{r}
#| message: false
#| warning: false

# Load all packages
library(tidyverse)          # Core data wrangling + visualisation
library(tsibble)            # Tidy time series structure
library(fable)              # Forecasting
library(feasts)             # Time series exploration tools
library(readxl)             # Importing Excel files
```

Think of the install command as buying a tool and placing it in your garage **(your computer's library).** The `library()` function is like grabbing the tool from your garage and bringing it to your workbench **(your current R session)** so you can use it

::: callout-tip
If you see a message like *"there is no package called...",* it means you have not installed it yet. Go back to the installation step.
:::
